\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-20 doc/incertae_sedis.tex
line breaks after .

@@

\author{Benjamin D. Redelings and Mark T. Holder} maketitle

@@

The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build
a comprehensive supertree \citep[see][]{redelings2017supertree}
covering all of life. The approach is to combine information in
published phylogenies with a comprehensive taxonomy that supplies
taxon names. The supertree method has been used with the Open Tree of
Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but in
principle is general. The taxonomy is a crucial input for several
reasons: it covers a wider range of species, its list of names allows
for alignment of different phylogenetic estimates to each other, and
it provides names for clades. Only a small proportion of known species
have been included in a phylogenetic analysis, thus a taxonomy is
crucial for achieving comprehensive coverage of known taxa.
Phylogenetic estimates collected from the published literature often
use different names for the same species. Lists of synonyms in OTT
(and other taxonomies) allow data curators to ``align'' input
phylogenies to the taxonomy \citep[see][ for discussion of the
curation process in the Open Tree of Life project]{McTavishEtAt2015}.
This alignment is important for recognizing when two different
estimates refer to the same taxon. Biologists are familiar with names
for ``higher'' taxa (taxa above the species rank). Thus, naming clades
in the supertree makes the tree easier to navigate and use.

@@

Flags. The Open Tree’s reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flagssee https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
with the flagging system per se, so “incertae sedis” will be used here
to refer to all of the flags that denote taxa with uncertain
placement.. While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.

Container nodes. Some taxonomies have nodes with names like “Incertae
sedis (Bacteria)”. Such nodes are called containers. They indicate
that each child node is an incertae sedis child of the container
node's parent. The OTT retains container nodes with the was_container
flag, but seems to have already moved all of the container's children
to the container's parent. }

@@

We define rooted splits for each edge by noting that removing the edge
would produce two trees. The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups. The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root. Such a split may be
written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e). \] where the
$\bullet$ indicates the root. The include set is the set of leaves
that are descendants of the daughter node of the edge. The exclude set
is the set of leaves that are not descendants of that node. If no taxa
are \emph{incertae sedis}, then the exclude set for a node is just the
total tip set minus the include set for the node:
$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes. For example, in Figure \ref{fig
:One-name-can}b, the name $A$ can apply to nodes $x$ and $y$. Here the
name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
$D$ since it is \emph{incertae sedis}. The two nodes $x$ and $y$
display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
B\,C$ respectively, and both of these splits display the split
$A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both $x$ and
$y$. This cannot happen without\emph{ incertae sedis} taxa except at
monotypic nodes. When faced with a choice about where to place a name,
our solution is to find the most tip-ward node where the name can
apply and attach the name to this node. This corresponds to a
conservative choice about whether \emph{incertae sedis} taxa should be
added to an existing taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}}
\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}}
\hfill{}
\caption{\label{fig:One-name-can}One name can be consistent with multiple nodes.}
\end{figure}

@@

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy. This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families. The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree. For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-
incertae-sedis}d), but not the taxonomy (Fig. \ref{fig:Handling-
incertae-sedis}b). Thus, we could say that the synthetic tree places
$C$ within $A$. Furthermore, the synthetic tree additionally places
$C$ within $AB$.

A slightly more general case is shown in Figure \ref{fig:Broken-
incertae-sedis}. Here taxon $C1$ is an descendant of $A$ in the
synthesis tree, but not the taxonomy. Thus $C1$ is placed within $A$.
We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
the taxonomy is the root node, and the path to the MRCA from $A$ is
$A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
root$. Here the additional node $AB$ on the path to the MRCA from $A$
indicates that $C1$ is placed within $AB$ as well as $A$. In contrast,
the additional node $C$ on the path to the MRCA indicates that $C1$ is
a broken \emph{incertae sedis} taxon. Thus we may consider an
algorithm that, for each taxon $B$ on the synthesis tree, finds the
closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
on the taxonomy. If not, we compute the paths $A\to A_{1}\to\ldots\to
MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

@@

When multiple names $N=\{n_{1},\ldots n_{N}\}$ map to the same
solution node $x$, then these names must satisfy some tree structure
on the taxonomy, such that $n_{1}<n_{2}$ if $n_{1}$ is a descendant of
$n_{2}$ in the taxonomy. If it is possible to find a name $n_{max}$
that is the unique maximal element of $N$, then it is permissible to
\begin{enumerate}
    \item create a monotypic parent $p(x)$ of $x$, and
assign $n_{max}$ to $p(x)$ 
    \item continue handling name clashes at $x$
with the set of possible names reduced to $N-n_{max}$. 
\end{enumerate}
However, its certainly possible that there might not be any such
$N_{max}$, in which case we could just choose a name for $x$ from $N$
(perhaps not an \emph{incertae sedis} name) and then record all the
other names as equivalents somewhere.

@@

The current conflict algorithm is pretty fast, but it works by
classifying tips into either (i) the include group or (ii) the exclude
group. To avoid counting the exclude group for every split, we instead
count the total number of children for each node, and assume that any
children not in the include set are in the exclude set. This is no
longer true when we have incertae sedis taxa. I suspect that if we
want to handle incertae sedies, we'd need a third category (iii) for
``neither include group nor exclude group''.
\begin{enumerate}
 \item
Get induced trees on intersection of leaf sets \item Compute depth for
each node (nd->depth)
 \item Compute number of tips at or below each
node (nd->n\_tips)
 \item for each input tree node -> $nd$
 \begin{enumerate}
   \item skip the root \item skip monotypic \item if
its a tip then find corresponding (``terminal'') edges in synth tree
and continue \item leaves1 <- get the list of leaves in the include
group of $nd$ (in input) \item L2 <- find the total number of tips (L2
= sum {[}nd->n\_tips| nd <- leaves1{]})
   \item leaves2 <- get list of
corresponding synth leaf nodes (in synth)
   \item nodes <- find all
nodes between leaves2 and the MRCA (in synth)
   \item MRCA <- mrca of
leaves2 (in synth. this uses the nd->depth annotation)
   \item Compute
number of tips in the include set (nd->include\_tips) below each node
in $nodes$ (in synth)
   \item if n\_include\_tips(MRCA) == n\_tips(MRCA)
then the MRCA displays $nd$
   \item if n\_include\_tips(MRCA) <
n\_tips(MRCA) then 
   \begin{itemize} 
     \item foreach node in nodes
      \begin{itemize}
         \item if (n\_include\_tips(nd) < n\_tips(nd) and
n\_include\_tips(nd) < l2)
          \begin{itemize}
            \item this is a conflict!
\end{itemize} 
           \end{itemize}
         \item if there are no conflicts, then this
is a resolved_by. 
      \end{itemize}
\end{enumerate}
\end{enumerate}

@@

This probably is outside the scope of the paper, but if we could come
up with a modified conflict algorithm, that would be nice/useful, and
probably novel. It would also probably be slower... 
\end{comment}

>>>>

% \raggedright

@@

\author{Benjamin D.~Redelings and Mark T.~Holder} 
\maketitle

@@

The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build
a comprehensive supertree \citep[see][]{redelings2017supertree}
covering all of life.
The approach is to combine information in
published phylogenies with a comprehensive taxonomy that supplies
taxon names.
The supertree method has been used with the Open Tree of
Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but in
principle is general.
The taxonomy is a crucial input for several
reasons: it covers a wider range of species, its list of names allows
for alignment of different phylogenetic estimates to each other, and
it provides names for clades.
Only a small proportion of known species
have been included in a phylogenetic analysis, thus a taxonomy is
important for achieving comprehensive coverage of known taxa.
Phylogenetic estimates collected from the published literature often
use different names for the same species.
Lists of synonyms in OTT
(and other taxonomies) allow data curators to ``align'' input
phylogenies to the taxonomy \citep[see][ for discussion of the
curation process in the Open Tree of Life project]{McTavishEtAt2015}.
This alignment is important for recognizing when two different
estimates refer to the same taxon.
Biologists are familiar with names
for ``higher'' taxa (taxa above the species rank).
Thus, naming clades
in the supertree makes the tree easier to navigate and use.

@@

Flags.
The Open Tree’s reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flagssee https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags.This document is not concerned
with the flagging system per se, so “incertae sedis” will be used here
to refer to all of the flags that denote taxa with uncertain
placement..
While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.
Container nodes.
Some taxonomies have nodes with names like “Incertae
sedis (Bacteria)”.
Such nodes are called containers.
They indicate
that each child node is an incertae sedis child of the container
node's parent.
The OTT retains container nodes with the was_container
flag, but seems to have already moved all of the container's children
to the container's parent.
}

@@

We define rooted splits for each edge by noting that removing the edge
would produce two trees.
The leaf label sets of these trees the
partition the full leaf set $\mathcal{L}$ into two groups.
The include
set $\mathcal{I}(e)$ which does not contain the root, and the exclude
set $\mathcal{E}(e)$ which does contain the root.
Such a split may be
written \[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\] where the
$\bullet$ indicates the root.
The include set is the set of leaves
that are descendants of the daughter node of the edge.
The exclude set
is the set of leaves that are not descendants of that node.
If no taxa
are \emph{incertae sedis}, then the exclude set for a node is just the
total tip set minus the include set for the node:
$\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$.

@@

However, in the \emph{incertae sedis} framework, it is possible for
one name to apply to multiple nodes.
For example, in Figure \ref{fig:One-name-can}b, 
the name $A$ can apply to nodes $x$ and $y$.
Here the
name $A$ corresponds to the split $A1\,A2\,|\bullet B\,C$, leaving out
$D$ since it is \emph{incertae sedis}.
The two nodes $x$ and $y$
display the splits $A1\,A2\,|\bullet B\,C\,D$ and $A1\,A2\,D|\bullet
B\,C$ respectively, and both of these splits display the split
$A1\,A2\,|\bullet B\,C$, so the name $A$ can apply to both $x$ and
$y$.
This cannot happen without\emph{ incertae sedis} taxa except at
monotypic nodes.
When faced with a choice about where to place a name,
our solution is to find the most tip-ward node where the name can
apply and attach the name to this node.
This corresponds to a
conservative choice about whether \emph{incertae sedis} taxa should be
added to an existing taxon.
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/tax-shaded}}
\hfill{}\subfloat[Solution tree]{\includegraphics[scale=0.5]{Figures/name1nodes2/solution-shaded}}
\hfill{}
\caption{\label{fig:One-name-can}One name can be consistent with multiple nodes.}
\end{figure}

@@

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy.
This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families.
The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree.
For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree 
(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
(Fig.\ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$. Furthermore, the synthetic tree additionally places
$C$ within $AB$.

A slightly more general case is shown in 
Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $C1$ is an descendant of $A$ in the
synthesis tree, but not the taxonomy.
Thus $C1$ is placed within $A$.
We note that the most recent common ancestor (MRCA) of $A$ and $C1$ in
the taxonomy is the root node, and the path to the MRCA from $A$ is
$A\to AB\to root$, while the path to the MRCA from $C1$ is $C1\to C\to
root$.
Here the additional node $AB$ on the path to the MRCA from $A$
indicates that $C1$ is placed within $AB$ as well as $A$.
In contrast,
the additional node $C$ on the path to the MRCA indicates that $C1$ is
a broken \emph{incertae sedis} taxon.
Thus we may consider an
algorithm that, for each taxon $B$ on the synthesis tree, finds the
closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
on the taxonomy.
If not, we compute the paths $A\to A_{1}\to\ldots\to
MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

@@

When multiple names $N=\{n_{1},\ldots n_{N}\}$ map to the same
solution node $x$, then these names must satisfy some tree structure
on the taxonomy, such that $n_{1}<n_{2}$ if $n_{1}$ is a descendant of
$n_{2}$ in the taxonomy.
If it is possible to find a name $n_{max}$
that is the unique maximal element of $N$, then it is permissible to
\begin{enumerate}
    \item create a monotypic parent $p(x)$ of $x$, and
assign $n_{max}$ to $p(x)$ 
    \item continue handling name clashes at $x$
with the set of possible names reduced to $N-n_{max}$.

\end{enumerate}
However, its certainly possible that there might not be any such
$N_{max}$, in which case we could just choose a name for $x$ from $N$
(perhaps not an \emph{incertae sedis} name) and then record all the
other names as equivalents somewhere.

@@

The current conflict algorithm is pretty fast, but it works by
classifying tips into either (i) the include group or (ii) the exclude
group.
To avoid counting the exclude group for every split, we instead
count the total number of children for each node, and assume that any
children not in the include set are in the exclude set.
This is no
longer true when we have incertae sedis taxa.
I suspect that if we
want to handle incertae sedies, we'd need a third category (iii) for
``neither include group nor exclude group''.
\begin{enumerate}
 \item
Get induced trees on intersection of leaf sets \item Compute depth for
each node (nd->depth)
 \item Compute number of tips at or below each
node (nd->n\_tips)
 \item for each input tree node -> $nd$
 \begin{enumerate}
   \item skip the root \item skip monotypic \item if
its a tip then find corresponding (``terminal'') edges in synth tree
and continue \item leaves1 <- get the list of leaves in the include
group of $nd$ (in input) \item L2 <- find the total number of tips (L2
= sum {[}nd->n\_tips| nd <- leaves1{]})
   \item leaves2 <- get list of
corresponding synth leaf nodes (in synth)
   \item nodes <- find all
nodes between leaves2 and the MRCA (in synth)
   \item MRCA <- mrca of
leaves2 (in synth.
this uses the nd->depth annotation)
   \item Compute
number of tips in the include set (nd->include\_tips) below each node
in $nodes$ (in synth)
   \item if n\_include\_tips(MRCA) == n\_tips(MRCA)
then the MRCA displays $nd$
   \item if n\_include\_tips(MRCA) <
n\_tips(MRCA) then 
   \begin{itemize} 
     \item foreach node in nodes
      \begin{itemize}
         \item if (n\_include\_tips(nd) < n\_tips(nd) and
n\_include\_tips(nd) < l2)
          \begin{itemize}
            \item this is a conflict!
\end{itemize} 
           \end{itemize}
         \item if there are no conflicts, then this
is a resolved_by.

      \end{itemize}
\end{enumerate}
\end{enumerate}

@@

This probably is outside the scope of the paper, but if we could come
up with a modified conflict algorithm, that would be nice/useful, and
probably novel.
It would also probably be slower...

\end{comment}
